Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections

### 2022-02-11 _paper.rmd
figure 8 caption

(ref:casefifa) FIFA 2020 data, a random forest model, regresses wages [2020 Euros] from nine aggregated skill measurements. The PI is a star offensive player (L. Messi) compared with a top defensive player (V. van Dijk). We remove three features with low attribution from both players. The attribution projection starts with the selected instance on the right. We vary the contribution from defense (`def`). The star offensive player is not distinguished in the horizontal direction. At this point, defensive players have been rotated to the highest horizontal value. (A video of the animated radial tour can be found at [vimeo.com/666431163](https://vimeo.com/666431163).)
```{r casefifa, echo=F, out.width = "100%", fig.cap = "(ref:casefifa)"}
knitr::include_graphics("./figures/case_fifa.png")
```

@@

Ames 2018, housing data was subset to North Ames (the neighborhood with the most house sales). The remaining are 338 house sales. A random forest model has regressed this price with the features Lot Area (`LtA`), Overall Quality (`Qlt`), Year the house was Built (`YrB`), Living Area (`LvA`), number of Bathrooms (`Bth`), number of Bedrooms (`Bdr`), total number of Rooms (`Rms`), Year the Garage was Built (`GYB`), and Garage Area (`GrA`). Using interaction from the global view, we select a house with an extreme negative residual and an accurate instance close to it in the data.

(ref:caseames) Exploring an instance with a large residual as the PI from fitting sales price [USD] to other variables in the Ames housing 2018. The sale price of the PI was under-predicted. The local explanation indicates a sizable attribution to Lot Area (`LtA`). The CI has a similar predicted sales price and smaller residual and has minimal attribution to Lot Area. In the attribution projection, the PI has a higher sales price than the CI. Reducing the contribution of lot area brings these two prices in line. This suggests if the model did not consider Lot Area, then the two houses would be quite similar. That is, the large residual is due to a lack of factoring in the lot area for the prediction of PI's sales price. (A video showing the animation is at [vimeo.com/666431134](https://vimeo.com/666431134).)
```{r caseames, echo=F, out.width="100%", fig.cap = "(ref:caseames)"}
knitr::include_graphics("./figures/case_ames2018.png")
```

>>>>

(ref:casefifa) Exploring the wages (euros) relative to skill measurements in the FIFA 2020 data. Star offensive player (L. Messi) is the PI and he is compared with a top defensive player (V. van Dijk). The attribution projection is shown at left, and it can  be seen that this combination of variables produces a view where Messi has very high predicted (and observed) wages. Defense (`def`) is the chosen variable to vary. It starts off very low, and as its contribution is increased (right plot) it can be seen that Messi's predicted wages decrease dramatically. The increased contribution in defense comes at the expense of offensive and reaction skills. The interpretation is that indeed, Messi's high wages are most attributable to his offensive and reaction skills, as initially provided by the local explanation. (A video of the animated radial tour can be found at [vimeo.com/666431163](https://vimeo.com/666431163).)
```{r casefifa, echo=F, out.width = "100%", fig.cap = "(ref:casefifa)"}
knitr::include_graphics("./figures/case_fifa.png")
```

@@

Ames 2018, housing data was subset to North Ames (the neighborhood with the most house sales). The remaining are 338 house sales. A random forest model has fitted price (USD) to the property features: Lot Area (`LtA`), Overall Quality (`Qlt`), Year the house was Built (`YrB`), Living Area (`LvA`), number of Bathrooms (`Bth`), number of Bedrooms (`Bdr`), total number of Rooms (`Rms`), Year the Garage was Built (`GYB`), and Garage Area (`GrA`). Using interaction from the global view, we select a house with an extreme negative residual and an accurate instance close to it in the data.

(ref:caseames) Exploring an instance with a large residual as the PI from fitting sales price (USD) to other variables in the Ames housing 2018. The sale price of the PI was under-predicted. The local explanation indicates a sizable attribution to Lot Area (`LtA`). The CI has a similar predicted sales price and smaller residual and has minimal attribution to Lot Area. In the attribution projection, the PI has a higher sales price than the CI. Reducing the contribution of lot area brings these two prices in line. This suggests if the model did not consider Lot Area, then the two houses would be quite similar. That is, the large residual is due to a lack of factoring in the lot area for the prediction of PI's sales price. (A video showing the animation is at [vimeo.com/666431134](https://vimeo.com/666431134).)
```{r caseames, echo=F, out.width="100%", fig.cap = "(ref:caseames)"}
knitr::include_graphics("./figures/case_ames2018.png")
```

### 2022-02-11 _paper.tex
figure 8 caption

\caption{FIFA 2020 data, a random forest model, regresses wages {[}2020 Euros{]} from nine aggregated skill measurements. The PI is a star offensive player (L. Messi) compared with a top defensive player (V. van Dijk). We remove three features with low attribution from both players. The attribution projection starts with the selected instance on the right. We vary the contribution from defense (\texttt{def}). The star offensive player is not distinguished in the horizontal direction. At this point, defensive players have been rotated to the highest horizontal value. (A video of the animated radial tour can be found at \href{https://vimeo.com/666431163}{vimeo.com/666431163}.)}\label{fig:casefifa}
\end{figure}

@@

Ames 2018, housing data was subset to North Ames (the neighborhood with the most house sales). The remaining are 338 house sales. A random forest model has regressed this price with the features Lot Area (\texttt{LtA}), Overall Quality (\texttt{Qlt}), Year the house was Built (\texttt{YrB}), Living Area (\texttt{LvA}), number of Bathrooms (\texttt{Bth}), number of Bedrooms (\texttt{Bdr}), total number of Rooms (\texttt{Rms}), Year the Garage was Built (\texttt{GYB}), and Garage Area (\texttt{GrA}). Using interaction from the global view, we select a house with an extreme negative residual and an accurate instance close to it in the data.

@@

\caption{Exploring an instance with a large residual as the PI from fitting sales price {[}USD{]} to other variables in the Ames housing 2018. The sale price of the PI was under-predicted. The local explanation indicates a sizable attribution to Lot Area (\texttt{LtA}). The CI has a similar predicted sales price and smaller residual and has minimal attribution to Lot Area. In the attribution projection, the PI has a higher sales price than the CI. Reducing the contribution of lot area brings these two prices in line. This suggests if the model did not consider Lot Area, then the two houses would be quite similar. That is, the large residual is due to a lack of factoring in the lot area for the prediction of PI's sales price. (A video showing the animation is at \href{https://vimeo.com/666431134}{vimeo.com/666431134}.)}\label{fig:caseames}
\end{figure}

>>>>

\caption{Exploring the wages (euros) relative to skill measurements in the FIFA 2020 data. Star offensive player (L. Messi) is the PI and he is compared with a top defensive player (V. van Dijk). The attribution projection is shown at left, and it can be seen that this combination of variables produces a view where Messi has very high predicted (and observed) wages. Defense (\texttt{def}) is the chosen variable to vary. It starts off very low, and as its contribution is increased (right plot) it can be seen that Messi's predicted wages decrease dramatically. The increased contribution in defense comes at the expense of offensive and reaction skills. The interpretation is that indeed, Messi's high wages are most attributable to his offensive and reaction skills, as initially provided by the local explanation. (A video of the animated radial tour can be found at \href{https://vimeo.com/666431163}{vimeo.com/666431163}.)}\label{fig:casefifa}
\end{figure}

@@

Ames 2018, housing data was subset to North Ames (the neighborhood with the most house sales). The remaining are 338 house sales. A random forest model has fitted price (USD) to the property features: Lot Area (\texttt{LtA}), Overall Quality (\texttt{Qlt}), Year the house was Built (\texttt{YrB}), Living Area (\texttt{LvA}), number of Bathrooms (\texttt{Bth}), number of Bedrooms (\texttt{Bdr}), total number of Rooms (\texttt{Rms}), Year the Garage was Built (\texttt{GYB}), and Garage Area (\texttt{GrA}). Using interaction from the global view, we select a house with an extreme negative residual and an accurate instance close to it in the data.

@@

\caption{Exploring an instance with a large residual as the PI from fitting sales price (USD) to other variables in the Ames housing 2018. The sale price of the PI was under-predicted. The local explanation indicates a sizable attribution to Lot Area (\texttt{LtA}). The CI has a similar predicted sales price and smaller residual and has minimal attribution to Lot Area. In the attribution projection, the PI has a higher sales price than the CI. Reducing the contribution of lot area brings these two prices in line. This suggests if the model did not consider Lot Area, then the two houses would be quite similar. That is, the large residual is due to a lack of factoring in the lot area for the prediction of PI's sales price. (A video showing the animation is at \href{https://vimeo.com/666431134}{vimeo.com/666431134}.)}\label{fig:caseames}
\end{figure}
